---
title: 'P&S-2022: Lab assignment 3'
author: "Pavlo Kryven, Fedir Zhurba, Ivan Nikolaichenko"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 80
---

# Part 1

# Markov chain

```{r}
TLN <- 21
set.seed(001)
SAMPLE_SIZE <- 1000
```

## Problem 1

### Estimate $\hat p$ that a random digit sequence of length $n = 100, 200, 1000$ contains the $TLN$

```{r}
generate_seq <- function(n){
  seq <- sample(0:9, n, replace = T)
  return (seq)
}
simulate_prob <- function(iters, n){
  prob_hat <- 0
  for(i in 1:iters){
    seq <- generate_seq(n)
    for (el in 1:(n-2)){
      if (seq[el] == 0 & seq[el+1]*10+seq[el+2] == TLN){
        prob_hat <- prob_hat + 1
        break
      } 
    }
  }
  return (prob_hat/(iters))
}
simulation_small <- simulate_prob(1000, 100)
simulation_medium <- simulate_prob(1000, 200)
simulation_large <- simulate_prob(1000, 1000)
simulation_very_large <- simulate_prob(1000, 10000)
print("For n = 100: ")
print(simulation_small)
print("For n = 200: ")
print(simulation_medium)
print("For n = 1000: ")
print(simulation_large)
print("For n = 10000: ")
print(simulation_very_large)

```

### Identify the Markov chain structure

So obviously we have 4 states corresponding to number of digits that matches in
our TLN. The transition probabilities matrix for our chain will look like this:
$$
    T :=
    \begin{pmatrix}
        0.9 & 0.1 & 0 & 0 \\
        0.8 & 0.1 & 0.1 & 0 \\
        0.8 & 0.1 & 0 & 0.1 \\
        0 & 0 & 0 & 1\\
    \end{pmatrix}
$$ It's based on the probability to get certain digit(we consider next digit
have uniform distribution). After that, we can solve corresponding system of
equations for limiting probabilities: $$
\begin{equation}
    \begin{cases}
      \pi_0 = 0.9*\pi_0+0.1*\pi_1\\
      \pi_1 = 0.8*\pi_0+0.1*\pi_1+0.1*\pi_2\\
      \pi_2 = 0.8*\pi_0+0.1*\pi_1+0.1*\pi_3\\
      \pi_3 = 0.9*\pi_0+0.1*\pi_1\\
      \pi_0 +\pi_1+\pi_2+\pi_3 = 1\\
    \end{cases}\,
\end{equation}
$$ From this system we get that $\pi_0 =\pi_1=\pi_2=0$ $\pi_3 = 1$. Indeed, this
is a logical result, as we have non-zero probability to go in $S_3$ and $S_3$ is
an absorbing state. So the limiting probability for this state is 1. This
theoretical probability is followed by our simulated $\hat p$, because as we can
see, with n becoming larger and larger, our estimated probability approaches 1.

### Determine approximately the sample size $N$ which guarantees the absolute error \|$\hat p_n$ − $p_n$\| of the estimate $\hat p_n$ is below 0.03 with confidence level of at least 95 percent

Using CLT, we can use our normal distribution of $\mu_{M_n}$ to calculate
Z-scores of sampling distribution. From there we will estimate confidence
intervals for true value of $\mu$, and other information. $$
P(|\hat p - p^*|\le\epsilon)\ge0.95 =>
P(|\hat p - \mu|\le\epsilon)\ge0.95 =>
P(|\hat p - \mu_{S_n}|\le\epsilon)\ge0.95
$$

The unbiased estimator of the probability that a random digit sequence of length
n contains the TLN is the proportion of sequences that contain the TLN.

$$ {p_n} = \frac{1}{N} \sum_{i=1}^N I(d_i \ldots d_n \text{ contains TLN}) $$

The variance of the estimator is:

$$ \text{SE}({p_n}) = \frac{s}{\sqrt{N}} $$

Confidence interval for the probability $p_n$ is defined as:

$$ 1 - \alpha = P(p_n \in \hat{p_n} \pm z_{\alpha/2} \sqrt{\frac{SE}{N}}) $$

```{r}
generate_n_probs <- function(iters, n, size){
  sampling_distr <- replicate(size, 0)
  for (i in 1:size){
    sampling_distr[i] <- simulate_prob(iters, n)
  }
  sampling_distr
}
SAMPLE_SIZE <- 1000
PROBS_SAMPLES <- 100
probs_distr <- generate_n_probs(100, SAMPLE_SIZE, PROBS_SAMPLES)
print("Distribution of probabilities: ")
print(probs_distr)
probs_mean <- mean(probs_distr)
print("Estimated probability(mean of simulations)")
print(probs_mean)

probs_std_dev <- sqrt(sum((probs_distr - probs_mean)^2)/(PROBS_SAMPLES-1))
probs_std_error <- probs_std_dev/sqrt(SAMPLE_SIZE)
print(probs_std_error)

z <- qnorm(1 - 0.05/ 2)
N <- ceiling(probs_mean * (1 - probs_mean) / (z^2 * 0.03^2))
print(N)
print("Confidence interval:")
print(c(probs_mean-z*probs_std_error, probs_mean+z*probs_std_error))
```

## Problem 2

### Make necessary amendments to the transition probabilities matrix P above and solve the corresponding system to find the expected time E(T) till absorption.

We can solve corresponding system of equations for expectation: $$
\begin{equation}
    \begin{cases}
      \mu_0 = 1+ 0.9*\mu_0+0.1*\mu_1\\
      \mu_1 = 1+ 0.8*\mu_0+0.1*\mu_1+0.1*\mu_2\\
      \mu_2 = 1+ 0.8*\mu_0+0.1*\mu_1+0.1*\mu_3\\
      \mu_3 = 0\\
    \end{cases}\,
\end{equation}
$$ We find that $\mu_0 = 1000$, $\mu_1 = 990$, $\mu_2 = 900$ \### Estimate
numerically the expected length E(T) till the first occurrence of the TLN by
running a sufficiently large number N of experiments.

```{r}
simulate_runs <- function(iters, n){
  expected_time <- 0
  for (i in 1:iters){
    seq <- generate_seq(n)
    time<- 3
    for (el in 1:(n-2)){
      
      if (seq[el] == 0 & seq[el+1]*10+seq[el+2] == TLN){
        expected_time = expected_time + time
        break
      } 
      else{
        time = time + 1
      }
    }
  }
   return (expected_time/iters)
}
runs_predicted <- simulate_runs(2000, 10000)
print("Estimator for E(T) is mean value amongst 1000 simulations")
print("Predicted expected time untill first TLN(021) occurance(mu_0):")
print(runs_predicted)

```

### Find the sample size N which guarantees that the absolute error $|\hat \theta - \theta|$ of the estimate does not exceed 10 with confidence level of at least 95 percent.

```{r}
PROBS_SAMPLES <- 100
generate_runs_distr <- function(iters, n, size){
  sampling_distr <- replicate(size, 0)
  for (i in 1:size){
    sampling_distr[i] <- simulate_runs(iters, n)
  }
  sampling_distr
}

runs_distr <- generate_runs_distr(300, 5000, PROBS_SAMPLES)
runs_std_dev <- sqrt(sum((runs_distr - 1000)^2)/(PROBS_SAMPLES-1))
runs_std_error <- runs_std_dev/sqrt(runs_predicted)
runs_var <- runs_std_error^2
print("Variance of runs length:")
print(runs_var)
```

L - length of run until TLN $P(|L - E(L)|\ge k)\le \frac{Var(L)}{10^2}$ by
Chebyshev's inequality $P(|L - E(L)|\le 10)\ge 1-\frac{Var(L)}{10^2}$ From here
we know that $\frac{Var(L)}{10^2} \le 0.05$ to satisfy the probability of at
least 95%. $Var(L)\le 5$ As we have that for N = 100\*300 (PROBS_SAMPLES times
iters) $Var(L)\approx 5$ that means that our searched $N \approx 30000$
